Quantifying the home ground advantage

### 2022-04-21 inst/home_advantage.Rmd
more work

(takes about ten minutes to run without cache)

Here I analyse a small synthetic dataset for home ground advantage.

@@

We see strong evidence to support the contention that home advantage is real.
Visually:

@@

>>>>

(takes about fifteen minutes to run without cache)

Here I analyse a small synthetic dataset for home ground advantage, then go on to discuss
a dataset presented by Agresti.

@@

We see strong evidence to support the contention that home advantage
is real.  Further, we may test the hypothesis that all the teams have
the same strength, after accounting for the home team advantage:

```{r,label=testhomegroundadvantageAgresti,cache=TRUE}
samep.test(H,teams)
```

Above, we see no evidence for a difference in team strengths.  Visually:

@@

# Example in Agresti

Agresti considers a dataset of seven baseball teams who play one
another repeatedly.  Each game is played at the home ground of one
team, the other playing away.  The dataset is as follows:

```{r,label=makeAgrestimatrix}
baseball_table <- matrix(c(
NA,   4+3i, 4+2i, 4+3i, 6+1i, 4+2i, 6+0i,
3+3i, NA  , 4+2i, 4+3i, 6+0i, 6+1i, 4+3i,
2+5i, 4+3i, NA  , 2+4i, 4+3i, 4+2i, 6+0i,
3+3i, 5+1i, 2+5i, NA  , 4+3i, 4+2i, 6+1i,
5+1i, 2+5i, 3+3i, 4+2i, NA  , 5+2i, 6+0i,
2+5i, 3+3i, 3+4i, 4+3i, 4+2i, NA  , 2+4i,
2+5i, 1+5i, 1+6i, 2+4i, 1+6i, 3+4i, NA),7,7)
teams <- c("Miulwaukee", "Detroit", "Toronto", "New York", "Boston","Cleveland","Baltimore")
rownames(baseball_table) <- teams
colnames(baseball_table) <- teams
M <- baseball_table # saves typing
```

Above, we represent home team wins with the real part of the matrix,
the imaginary component being away wins.  Now process it:

```{r,label=makesuppfrommatrix}
baseball <- home_away(M)
baseball
```

Maximum likelihood:

```{r,label=calculatemaxpAgresti,cache=TRUE}
baseball_maxp <- maxp(baseball)
baseball_maxp
```

Visually:

```{r,label=showmaxpAgresti}
dotchart(baseball_maxp,pch=16)
pie(baseball_maxp)
```

Above, we see that the teams are all of roughly equal strength, and
the home advantage is small.  However, we may easily assess the null
that the home strength is zero:

```{r,analyseAgresti,cache=TRUE}
specificp.gt.test(baseball,"home",0)
```

```{r,Agrestiprofsupp,cache=TRUE}
home_strength <- seq(from=0.005,to=0.1,length=20)
plot(home_strength,profsupp(baseball,"home",home_strength),type="b",pch=16,
     xlab="home strength",ylab="support",main="profile likelihood for home advantage")
abline(h=c(0,-2))
```

Further, we may test the hypothesis that the baseball teams have the same strength:

```{r,label=testhomegroundadvantage,cache=TRUE}
samep.test(baseball,teams)
```

Above, we see no evidence for the teams having differential strengths.

### Some further thoughts on the home ground monster and Agresti's analysis.

Considering the simplest case of two teams, a and b, with results:

* $a$ at home, $b$ away, $a,b$ wins with probability $\frac{\lambda a,b}{\lambda a+b}$ (Agresti), or $\frac{a+H,b}{a+b+H}$ (me). 
* $a$ away, $b$ at home, $a,b$ wins with probability $\frac{a,\lambda b}{\lambda a+b}$ (Agresti), or $\frac{a,b+H}{a+b+H}$ (me).

Matching the probabilities for $a$ at home gives us $H=a(\lambda-1)$
and matching the probabilities for $a$ away gives $H=b(\lambda-1)$.
These cannot both be true unless $a=b$ or $\lambda=1$.  But we can, in principle,
test which one is a better model.

### Package dataset {-}

Following lines create `home_advantage.rda`, residing in the `data/`
directory of the package.

```{r,label=savesolingdataset}
save(baseball_table, baseball, baseball_maxp, 
  file="baseball.rda")
```
